Inferring Galactic Parameters from Chemical Abundances with Simulation-Based Inference

### 2024-12-17 src/tex/ms.tex
[showyourwork] overleaf sync

The exact values of parameters sensitively affect chemical evolution tracks \citep{2005A&A...430..491R,2015MNRAS.449.1327V,2015MNRAS.451.3693M}, yet their observational constraints remain weak.
%
For example, a range of high-mass IMF slopes have been suggested \citep[Tab.\,7]{2016ApJ...824...82C}, with a steeper-than-canonical slope being suggested by a range of studies \citep[e.g.][]{2015ApJ...806..198W,2015MNRAS.447.3880R,2014ApJ...796...75C}. In addition, the IMF slope may itself be not a constant but rather a function of metallicity, introducing further complexity \citep[e.g.][]{2019MNRAS.482..118G,2019A&A...626A.124M}. Similarly, the choice of SN\,Ia delay-time-distribution and normalization plays a crucial role in the enrichment of the ISM \citep[e.g.][]{Buck2021} and is heavily debated \citep{2010ApJ...722.1879M,2012MNRAS.426.3282M,2015ApJ...810..137J}.

The goal of this work is to demonstrate how we can use modern machine learning techniques in tandem with a simple galactic chemical evolution (GCE) model in a Bayesian framework to infer global galactic parameters from a set of stars. Our work focuses on two key parameters; the high-mass slope of the \citet[Tab.\,1]{2003PASP..115..763C} IMF and the rate of SN\,Ia explosions per unit mass. We assume that both are constant across the galaxy and time.

Our method will make use of simulation-based inference \citep[SBI, e.g.][]{Cranmer2020}. For high-dimensional posterior functions, SBI gives tremendous computational improvements over more traditional methods such as Markov Chain Monte Carlo (MCMC) or even Hamilton Monte Carlo (HMC) methods.

\textcolor{red}{to be adapted...\\
This paper is structured as follows. In Section \ref{sec:methods} we detail our methods. Especially, Section \ref{sec:dataset_generation} discusses the methodology used for generating the dataset for our analysis by extracting galaxy data from simulations. Section \ref{sec: methods_PCA_on_images} outlines the application of PCA to the image data and the resulting lower-dimensional image space (Section \ref{sec: dimensionality reduction method using pca}). Our resulting model is presented in Section \ref{sec: Results} and its accuracy thoroughly tested in Section \ref{sec: model evaluation, reconstruction error}. In Section \ref{sec: application} we highlight potential applications of our PCA galaxy model and conclude in Section \ref{sec: conclusion} with a summary and outlook of our results.}

@@

\paragraph{Galactic chemical evolution models}
Our simulator is based on the \texttt{CHEMPYScoring} module \citep{Philcox_2018} publicly available as the \texttt{CHEMPYMulti} \citep{Philcox_2019}\footnote{\href{https://github.com/oliverphilcox/ChempyMulti}{github.com/oliverphilcox/ChempyMulti}} package a further development of the original \texttt{CHEMPY} model \citep{Rybizki_2017}. \texttt{CHEMPY} is a simple Galactic Chemical Evolution (GCE) model that is able to predict stellar chemical abundances throughout cosmic time by using published nucleosynthetic yield tables for three key processes (SN\,Ia and SN\,II explosions and AGB stellar feedback) and a small number of parameters controlling simple stellar populations (SSPs) and ISM physics. We refer to the initial \texttt{CHEMPY} paper \citep{Rybizki_2017} for the details of the model.
% %
In this work, we allow six \texttt{CHEMPY} parameters to vary freely (see also Tab.\,\ref{tab:priors}). These can be categorized into three groups:
\vspace{-1em}
% %\setlist[1]{itemsep=-4pt}
\begin{enumerate}
     \item $\vec\Lambda$: \textbf{Global Galactic Parameters} describe SSP physics and comprise the high-mass \citet{2003PASP..115..763C} IMF slope, $\alpha_\mathrm{IMF}$, and (logarithmic) Type Ia SN normalization, $\log_{10}(N_\mathrm{Ia})$. We treat these as star-independent and assume them to be constant across galactic environments and cosmic time. \footnote{Whilst $\log_{10}(N_\mathrm{Ia})$ is constant with respect to time by definition, it being simply a normalization constant, there is some evidence for $\alpha_\mathrm{IMF}$ varying as a function of time or metallicity \citep{2014ApJ...796...75C,2016MNRAS.462.2832C,2019MNRAS.482..118G,Martin2019}.} 
     We adopt the same broad priors as \citep{Philcox_2019} for these variables (see also Tab.\ref{tab:priors}).  
     \item $\{\vec\Theta_i\}$: \textbf{Local Galactic Parameters} describe the local physics of the ISM and are hence specific to each stellar environment, indexed by $i$. As defined in \citep{Rybizki_2017}, these include the star-formation efficiency (SFE), $\log_{10}(\text{SFE})$, $\log_{10}(\mathrm{SFR}_\mathrm{peak})$, which controls the peak of the star formation rate (SFR), and the fraction of stellar outflow that is fed to the gas reservoir, $\mathrm{x}_\mathrm{out}$. We adopt broad priors for all parameters and, as in \citep{Philcox_2019}, fix the SN\,Ia delay-time distribution, $\log_{10}(\tau_{\rm Ia})$, to $\log_{10}(\tau_{\rm Ia})=-0.80$ \citep[see also][]{Philcox_2018}.
     \item $\{T_i\}$: \textbf{Stellar Birth-Times}. Time in Gyr at which a given star is formed from the ISM. We assume that its proto-stellar abundances match the local ISM abundances at $T_i$.
\end{enumerate}

The separability of local (ISM) parameters and global (SSP) parameters is motivated by recent observational evidence: \citet{2019arXiv190710606N} find that the elemental abundances of red clump stars belonging to the thin disk can be predicted almost perfectly from their age and [Fe/H] abundance. This implies that the key chemical evolution parameters affecting the elemental abundances (SSP parameters and yield tables) are held fixed, whilst ISM parameters vary smoothly over the thin disk (which offsets the metallicity for different galactocentric radii). Similarly \cite{2019ApJ...874..102W} find that ISM parameter variations are deprojected in the [X/Mg] vs [Mg/H] plane (their Fig.\,17) and that abundance tracks in that space are independent of the stellar sample's spatial position within the Galaxy (their Fig.\,3).

@@

We adopt the same nucleosynthetic yield tables as in \citep{Philcox_2019}, see their Sec.~2.2 for more details.
To ensure maximal compatibility with TNG, we adopt their nucleosynthetic yield tables in \texttt{Chempy}, for enrichment by SN\,Ia, SN\,II and AGB stars. The utilized yields are summarized in Tab.\,\ref{tab:chempy_TNG_yields}, matching \citet[Tab.\,2]{2018MNRAS.473.4077P}, and we note that the SN\,II yields are renormalized such that the IMF-weighted yield ratios at each metallicity are equal to those from the \citet{2006ApJ...653.1145K} mass range models alone. \texttt{Chempy} uses only net yields, such that they provide only newly synthesized material, with the remainder coming from the initial SSP composition. These tables may not well-represent true stellar chemistry, and the effects of this are examined in Sec.\,\ref{subsec: mocks_wrong_yield} by performing inference using an alternative set of yields. For the analysis of observational data, we would want to use the most up-to-date yields, such as \citet{2016ApJ...825...26K} AGB yields, and carefully chose elements which are known to be well reproduced by our current models (e.g. shown by \citet{2019ApJ...874..102W,2019arXiv190806113G}), though this is not appropriate in our context. To facilitate best comparison with TNG, we further set the maximum SN\,II mass as $100\,\mathrm{M}_\odot$ (matching the IMF upper mass limit), adopt stellar lifetimes from \citet{1998A&A...334..505P} and do not allow for any `hypernovae' (in contrary to \citetalias{2018ApJ...861...40P}).

@@

%TNG only tracks nine elements in their analysis: C, Fe, H, He, Mg, N, Ne, O and Si, reporting the mass-fractions of each \citep{2018MNRAS.473.4077P}. 
In our analysis we only track nine elements: C, Fe, H, He, Mg, N, Ne, O and Si and principally compare the logarithmic abundances [X/Fe] and [Fe/H] (defined by 
\begin{equation}
     [\mathrm{X}/\mathrm{Y}] = \log_{10}(N_\mathrm X/N_\mathrm Y)_\mathrm{star} - \log_{10}(N_\mathrm X/N_\mathrm Y)_\odot
\end{equation}
for number fraction $N_\mathrm X$ of element X), where $\odot$ denotes the solar number fractions of \citet{2009ARA&A..47..481A}. This uses H for normalization, thus we are left with $n_\mathrm{el}=8$ independent elements which must be tracked by \texttt{Chempy}.\footnote{In observational contexts, it may be more appropriate to compute abundances relative to Mg rather than Fe (as in \citep{2019ApJ...874..102W}) since Mg is only significantly produced by SN\,II and hence a simpler tracer of chemical enrichment.}

In this paper, \texttt{Chempy} will be used as the principal GCE model, which, with the modifications described above, allows for fast prediction of TNG-like chemical abundances for a given set of galactic parameters. It is important to note that the two GCE models have very different parametrizations of galactic physics, with TNG including vastly more effects, thus it is not certain \textit{a priori} how useful \texttt{Chempy} will be in emulating the TNG simulation, although its utility was partially demonstrated in \citet{2018ApJ...861...40P}. This is a necessary test to prepare for an inference on real data.

\paragraph{Neural network emulator for \texttt{CHEMPY}:}
Despite the simplifications made by the GCE model \texttt{CHEMPY}, we still have difficulties sampling the distribution of the global parameters $\vec\Lambda = \{\alpha_\mathrm{IMF},\log_{10}(N_\mathrm{Ia})\}$ due to the run-time of \texttt{CHEMPY} and the high-dimensionality of the parameter space. To alleviate this, we follow \cite{Philcox_2019} and implement a \textit{neural network} (NN) emulator of the \texttt{CHEMPY} simulations. The NN is a simple feed-forward neural network with 2 hidden layers and 100 neurons in the first and 40 neurons in the second layer.

In essence, instead of computing the full model for each input parameter set, we pass the parameters to the network which predicts the output abundances to high accuracy. As already argued in \cite{Philcox_2019} this has two benefits;
\begin{enumerate}
    \item \textbf{Speed:} The run-time of the \texttt{Chempy} function is $\sim1$\,s per input parameter set, which leads to very slow posterior sampling. With the neural network, this reduces to $\sim5\times10^{-5}$\,s, and is trivially parallelizable, unlike \texttt{Chempy}.
    \item \textbf{Differentiablility:} The neural network has a simple closed-form analytic structure (described in appendix \ref{appen: neural_network}), unlike the complex \texttt{Chempy} model. This allows it to be differentiated, so we can sample via advanced methods (cf.\,Sec.\,\ref{sec: sampling}).
\end{enumerate}

Despite the additional complexity introduced by using multiple stellar data-points, our network simply needs to predict the birth-time abundances for a single star (with index $i$) given a set of six parameters; $\{\vec\Lambda,\vec\Theta_i,T_i\}$. The same network can be used for all $n_\mathrm{stars}$ stars (and run in parallel), reducing a set of $n_\mathrm{stars}$ runs of \texttt{Chempy} to a single matrix computation. With the above choices, the network predicts abundances with an absolute percantage error of $1.6_{-0.9}^{+2.4}$\,\%, \textcolor{red}{is this still the right value? using showyourwork, we could do a little evaluation script to have always the right number :-)} \textcolor{blue}{Yep, also correct} which is far below typical observational errors and even smaller away from the extremes of parameter space.
All our code to reproduce our results is publicly avaialble on \href{https://github.com/bGuenes/sbi_chemical_abundances}{github}.

@@

>>>>

The exact values of parameters sensitively affect chemical evolution tracks \citep{2005A&A...430..491R,2015MNRAS.449.1327V,2015MNRAS.451.3693M}, yet their observational constraints remain weak.
%
For example, a range of high-mass IMF slopes have been suggested \citep[Tab.\,7]{2016ApJ...824...82C}, with a steeper-than-canonical slope being suggested by a range of studies \citep[e.g.][]{2015ApJ...806..198W,Rybizki2015,Chabrier2014}. In addition, the IMF slope may itself be not a constant but rather a function of metallicity, introducing further complexity \citep[e.g.][]{2019MNRAS.482..118G,Martin2019}. Similarly, the choice of SN\,Ia delay-time-distribution and normalization plays a crucial role in the enrichment of the ISM \citep[e.g.][]{Buck2021} and is heavily debated \citep{2010ApJ...722.1879M,2012MNRAS.426.3282M,2015ApJ...810..137J}.

The goal of this work is to demonstrate how we can use modern machine learning (ML) techniques in tandem with a simple galactic chemical evolution (GCE) model in a Bayesian framework to infer global galactic parameters from a set of stars. Our work focuses on two key parameters; the high-mass slope of the \citet[Tab.\,1]{2003PASP..115..763C} IMF and the rate of SN\,Ia explosions per unit mass. We assume that both are constant across the galaxy and time.

A methodologically very similar work was previously presented by \cite{Philcox_2019} using Hamilton Monte Carlo sampling for the inference. The result of this work was an impressive speed-up over classical Markov Chain Monte Carlo sampling with no loss of accuracy. Here we make use of simulation-based inference \citep[SBI, e.g.][]{Cranmer2020}. For high-dimensional posterior functions, SBI gives tremendous computational improvements over more traditional methods such as Markov Chain Monte Carlo (MCMC) or even the more efficient Hamilton Monte Carlo (HMC) methods.

This paper is structured as follows. In Section~\ref{sec:methods} we detail our methods. In particular, we discuss the GCE model and the methods behind SBI. Our results are presented in Section~\ref{sec: Results} where we test the performance of our approach on mock data and data taken from a hydrodynamical simulation. We conclude our study with Section~\ref{sec: conclusion} with a summary and outlook of our results.

@@

In order to establish our new method based on simulation based inference we need two ingredients: A simulator (in our case a Galactic chemical evolution model) that simulates observational data from a set of model parameters (in our case the IMF slope and the Type Ia supernovae normalization) and a flexible way of parametrizing the posterior density conditioned on the observation in order to perform our inference. In the next subsections we describe both ingredients in detail.

\subsection{Galactic chemical evolution models}
Our simulator is based on the \texttt{CHEMPY} model \citep{Rybizki_2017}. \texttt{CHEMPY} is a simple Galactic Chemical Evolution (GCE) model that is able to predict stellar chemical abundances throughout cosmic time by using published nucleosynthetic yield tables for three key processes (SN\,Ia and SN\,II explosions and AGB stellar feedback) and a small number of parameters controlling simple stellar populations (SSPs) and ISM physics. We refer to the initial \texttt{CHEMPY} paper \citep{Rybizki_2017} for the details of the model.

In particular, we are using the \texttt{CHEMPYScoring} module \citep{Philcox_2018} publicly available as the \texttt{CHEMPYMulti} \citep{Philcox_2019}\footnote{\href{https://github.com/oliverphilcox/ChempyMulti}{github.com/oliverphilcox/ChempyMulti}} package a further development of the original \texttt{CHEMPY} model.

\paragraph{\texttt{CHEMPY} parameters}
In this work, we allow six \texttt{CHEMPY} parameters to vary freely (see also Tab.\,\ref{tab:priors}). These can be categorized into three groups:
\begin{enumerate}
     \item $\vec\Lambda$: \textbf{Global Galactic Parameters} describe SSP physics and comprise the high-mass \citet{2003PASP..115..763C} IMF slope, $\alpha_\mathrm{IMF}$, and (logarithmic) Type Ia SN normalization, $\log_{10}(N_\mathrm{Ia})$. We treat these as star-independent and assume them to be constant across galactic environments and cosmic time\footnote{Whilst $\log_{10}(N_\mathrm{Ia})$ is constant with respect to time by definition, it being simply a normalization constant, there is some evidence for $\alpha_\mathrm{IMF}$ varying as a function of time or metallicity \citep{Chabrier2014,2016MNRAS.462.2832C,2019MNRAS.482..118G,Martin2019}.}. 
     We adopt the same broad priors as \citep{Philcox_2019} for these variables (see also Tab.\ref{tab:priors}). 
     %
     \item $\{\vec\Theta_i\}$: \textbf{Local Galactic Parameters} describe the local physics of the ISM and are hence specific to each stellar environment, indexed by $i$. As defined in \citep{Rybizki_2017}, these include the star-formation efficiency (SFE), $\log_{10}(\text{SFE})$, $\log_{10}(\mathrm{SFR}_\mathrm{peak})$, which controls the peak of the star formation rate (SFR), and the fraction of stellar outflow that is fed to the gas reservoir, $\mathrm{x}_\mathrm{out}$. We adopt broad priors for all parameters and, as in \citep{Philcox_2019}, fix the SN\,Ia delay-time distribution, $\log_{10}(\tau_{\rm Ia})$, to $\log_{10}(\tau_{\rm Ia})=-0.80$ \citep[see also][]{Philcox_2018}.
     \item $\{T_i\}$: \textbf{Stellar Birth-Times}. Time in Gyr at which a given star is formed from the ISM. We assume that its proto-stellar abundances match the local ISM abundances at $T_i$.
\end{enumerate}

The separability of local (ISM) parameters and global (SSP) parameters is motivated by recent observational evidence: \citet{2019arXiv190710606N} find that the elemental abundances of red clump stars belonging to the thin disk can be predicted almost perfectly from their age and [Fe/H] abundance. This implies that the key chemical evolution parameters affecting the elemental abundances (SSP parameters and yield tables) are held fixed, whilst ISM parameters vary smoothly over the thin disk \citep[which offsets the metallicity for different galactocentric radii, e.g.][for a simulated example]{Buck2020, Wang2024}. Similarly \cite{2019ApJ...874..102W} find that ISM parameter variations are deprojected in the [X/Mg] vs [Mg/H] plane (their Fig.\,17) and that abundance tracks in that space are independent of the stellar sample's spatial position within the Galaxy (their Fig.\,3).

@@

\paragraph{Nucleosynthetic yield tables}
We adopt the same nucleosynthetic yield tables as in \citep{Philcox_2019}, see their Sec.~2.2 for more details.
To test our method, we aim further at inferring parameters from a sample of stars taken from a hydrodynamical simulation of a MW type galaxy which we take from the Ilustris TNG project \citep{Pillepich2018}. To ensure maximal compatibility with TNG, we adopt their nucleosynthetic yield tables in \texttt{Chempy}, for enrichment by SN\,Ia, SN\,II and AGB stars. The utilized yields are summarized in Tab.\,\ref{tab:chempy_TNG_yields}, matching \citet[Tab.\,2]{2018MNRAS.473.4077P}, and we note that the SN\,II yields are renormalized such that the IMF-weighted yield ratios at each metallicity are equal to those from the \citet{2006ApJ...653.1145K} mass range models alone. \texttt{Chempy} uses only net yields, such that they provide only newly synthesized material, with the remainder coming from the initial SSP composition. These tables may not well-represent true stellar chemistry, and the effects of this missmatch are examined in Sec.\,\ref{subsec: mocks_wrong_yield} by performing inference using an alternative set of yields that does not match the yield set of the training data. For the analysis of observational data, we would want to use the most up-to-date yields, such as \citet{2016ApJ...825...26K} AGB yields, and carefully chose elements which are known to be well reproduced by our current models (e.g. shown by \citet{2019ApJ...874..102W,2019arXiv190806113G}), though this is not appropriate in our context. To facilitate best comparison with TNG, we further set the maximum SN\,II mass as $100\,\mathrm{M}_\odot$ (matching the IMF upper mass limit), adopt stellar lifetimes from \citet{1998A&A...334..505P} and do not allow for any `hypernovae' \citep[in contrary to][]{2018ApJ...861...40P}).

@@

\paragraph{Chemical elements}
In our analysis we only track nine elements: C, Fe, H, He, Mg, N, Ne, O and Si since these are the only elements traced by TNG. We principally compare the logarithmic abundances [X/Fe] and [Fe/H] defined by 
\begin{equation}
     [\mathrm{X}/\mathrm{Y}] = \log_{10}(N_\mathrm X/N_\mathrm Y)_\mathrm{star} - \log_{10}(N_\mathrm X/N_\mathrm Y)_\odot
\end{equation}
for number fraction $N_\mathrm X$ of element X. Here $\odot$ denotes the solar number fractions of \citet{2009ARA&A..47..481A}. This uses H for normalization, thus we are left with $n_\mathrm{el}=8$ independent elements which must be tracked by \texttt{Chempy}\footnote{In observational contexts, it may be more appropriate to compute abundances relative to Mg rather than Fe (as in \citep{2019ApJ...874..102W}) since Mg is only significantly produced by SN\,II and hence a simpler tracer of chemical enrichment.}.

With these modifications, \texttt{Chempy} allows for fast prediction of TNG-like chemical abundances for a given set of galactic parameters. It is important to note that the two GCE models have very different parametrizations of galactic physics, with TNG including vastly more effects, thus it is not certain \textit{a priori} how useful \texttt{Chempy} will be in emulating the TNG simulation, although its utility was partially demonstrated in \citet{2018ApJ...861...40P}. However,  such a test is necessary to prepare for an inference on real data.

\subsection{Neural network emulator for \texttt{CHEMPY}:}
Despite the simplifications made by the GCE model \texttt{CHEMPY}, the run-time of \texttt{CHEMPY} and the high-dimensionality of the parameter space incurs some difficulties when sampling the distribution of the global parameters $\vec\Lambda = \{\alpha_\mathrm{IMF},\log_{10}(N_\mathrm{Ia})\}$. 
To alleviate this, we follow \cite{Philcox_2019} and implement a \textit{neural network} (NN) emulator of the \texttt{CHEMPY} simulator. We design the NN as a simple feed-forward neural network with 2 hidden layers and 100 neurons in the first and 40 neurons in the second layer.

In essence, instead of computing the full model for each input parameter set, we pass the parameters to the NN which predicts the output abundances to high accuracy. As already argued in \cite{Philcox_2019} this has two benefits;
\begin{enumerate}
    \item \textbf{Speed:} The run-time of the \texttt{Chempy} function is $\sim1$\,s per input parameter set, which leads to very slow generation of training data for SBI. With the NN emulator, this reduces to $\sim5\times10^{-5}$\,s, and is trivially parallelizable, unlike \texttt{Chempy}.
    \item \textbf{Differentiability:} The NN is written in pytorch which allows for automatic differentiation. Additionally the NN has a simple closed-form analytic structure \citep[described in the appendix of][]{Philcox_2019}, unlike the complex \texttt{Chempy} model. This allows it to be differentiated, so one can use it to sample via advanced methods such HMC as done in \citep{Philcox_2019}.
\end{enumerate}

Despite the additional complexity introduced by using multiple stellar data-points, our NN simply needs to predict the birth-time abundances for a single star (with index $i$) given a set of six parameters; $\{\vec\Lambda,\vec\Theta_i,T_i\}$. The same NN can be used for all $n_\mathrm{stars}$ stars (and run in parallel), reducing a set of $n_\mathrm{stars}$ runs of \texttt{Chempy} to a single matrix computation. With the above network parameter choices, the NN predicts abundances with an absolute percantage error of $1.6_{-0.9}^{+2.4}$\,\%, which is far below typical observational errors and even smaller away from the extremes of parameter space.
Our NN emulator is publicly available on the github repository accompanying this manuscript \href{https://github.com/TobiBu/sbi-chempy}{https://github.com/TobiBu/sbi-chempy}. \textcolor{red}{adapt to point to final folder on the repo. and maybe add zenodo link for the nueral network weights...}

\begin{figure}[]
     \centering
     \includegraphics[width=\columnwidth]{figures/ape_NN.png}
     \vspace{-.5cm}
     \caption{Absolute percentage error (APE) of the NN emulator for the \texttt{CHEMPY} simulator.}
     \label{fig:ape_NN}
     \script{train_torch_chempy.py}
\end{figure}

@@

\subsection{Alternative yield set}
\label{subsec: mocks_wrong_yield}
